Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities

### 2023-07-16 text/sections/00-abstract.tex
writing revision 2 mostly finished.

US universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while at the same time states have limited how much they fund higher education.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall of $10$\% in funding for a public university, via a state funding shock, decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%; long-run estimates show that the faculty composition is affected both in the same year as the funding shock, and two to three years following.
Analysis of all the professors at Illinois public universities shows that incumbent professors are not affected by substantial funding shocks over 2011-2021, with suggestive evidence showing that the faculty changes arose by limiting the hiring of new professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

>>>>

US universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while states have limited how much they fund higher education.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall in funding of \$1,000 per student leads to univeristies employing 6 more lecturers on average; relative to enrolment, a fall of $10$\% in funding per student decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%.
Dynamic estimates show that state funding cuts affect composition of faculty for the next 3 years, showing that funding cuts have lingering effects.
Analysis of all the professors at Illinois public universities shows that incumbent professors' salaries, promotion rate, and quit rate are all are not affected by funding cuts over 2011-2021, implying that the composition change arose by limiting the hiring of new professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

### 2023-07-16 text/sections/01-introduction.tex
writing revision 2 mostly finished.

Public universities educate the majority of higher education students in the US, yet have experienced a stagnation in state funding over the last three decades .
At the same time, public university employment of lecturers dramatically increased at public universities, while enrolment rose dramatically.
Previous research has examined the effects of the decline in state funding on student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence of the effects of thid funding decline on faculty.
I show that four-year, degree-granting public universities have experienced a systematic decline in state funding per student, and use a shift-share instrument to show that this fall led to a substitution away from tenure-track/tenured professors towards contingent lecturers.
Analsysis of every public university faculty member in the state of Illinois shows that incumbent faculty were relatively unaffected, implying that changes in faculty composition arose by disrupting public universities' ability to hire or replace their tenure-track/tenured faculty. 
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors, so that the stagnating state support for public universities has implications for the wider structure of US higher education and research.

There are various margins through which state fundings cuts can impact a university's faculty.
A university's level of resources affect the ability that departments can hire new professors, retain their current fauclty, or provide retirement incentives.
So that it is natural to expect that funding cuts may affect the composition of professors at a university.
Similarly, the wages a university pays their faculty is affected by funding for the university, including the starting salary it offers to new professors and the amount of yearly raises (if any) for incumbent faculty.
So that it is natural to expect that systematic funding cuts will lead to changes in faculty composition, and possibly affect incumbent professors, too.
Yet, it is a priori unclear whether/how much each of these margins adjusts to funding cuts, as these decision depend on how the university administrations view the productivity of different inputs, plus the degree of wage and employment stickiness over time.

@@

I employ a shift-share instrument to identify changes in state funding for higher education, which exploits how universities differ in their financial reliance on state funding, combined with yearly shocks to the amount an entire state funds higher education --- following \cite{NBERw23736,NBERw27885}.
State governments decide how to fund higher education, and everything else, by a complicated process which can be influenced by local economic conditions, changing state priorities, or even lobbying from the state universities themselves.
As such, state funding is not randomly assigned to universities and so it is necessary to employ an identification strategy, such as the shift-share instrument, to account for this.
The shift-share instrument interacts reliance on state funding in a base period with the entire state's total funding per higher education student for each year, to exploit changes in funding and how much each university historically relied on state funding for higher education.

\textbf{Adjust this to have dollar figures, and generall re-do desc of results.}
Falls in state funding affects faculty composition at public universities, which substitute away from tenured and tenure-track professors towards contingent lecturers.
A fall of $-10$\% in state funding leads to a fall of $-1.4$\% in the number of assistant professors per student at a university, a fall of $-1.2$\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock, showing that the effect is not isolated to the immediate year of the first funding shock. 
Over the same time period, state funding fell by around $-$35\%, while the count of professors per student fell by $-$9\%, so that these results show that falls in state funding explain around a third of the substitution away from tenure-track and tenure professors towards contingent lecturers.
Additionally, I use individual level data on all Illinois public university professors in 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by the falls in state funding (and by the funding shocks).
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or leaving) professors.

\textit{Re-do the lit review to focus on mechanisms of student quality.}
A number of studies have shown that stagnating state funding for higher education negatively impacts student outcomes.
\cite{NBERw23736} show that increases in public university spending (via funding shocks) increases enrolment and degree completion among students,\footnote{
    \cite{miller2022making} further analyse the effects of falls in university revenues, by finding that reductions in public universities tuition prices (suggestively motivated by falling state funding) leads to reductions in provided financial aid.
}
that high school students exposed to similar public school funding shocks are less likely to attend college \citep{jackson2021school}, and that these shocks induce public universities to rely more on tuition as a source of revenue \citep{bound2019public}.
\textbf{Faculty outcomes surveyed by \cite{ehrenberg2003studying}.}
\cite{NBERw27885} use the same methodology to show that increases in state funding lead to degree lower completion time and later-life debt for students.\footnote{
    \cite{bound2007cohort} use similar variation in state funding per student to show that lower funding leads to lower completion rates at public universities.
}

Over the last half century, there have been drastic changes to the heirarchy of US higher education, where the top universities have become more selective for student applicants, and the average university less selective \citep{hoxby2009changing}.
Similarly, research faculty at US universities are increasingly concentrated among the graduates of highly selective PhD programmes \citep{wapman2022quantifying}.
Faculty composition is an interesting outcome on its own, but its relationship with higher education finances has so far not been studied.
The closest previous examples study  how university endowments react to negative endowment shocks \citep{brown2014endowment} and the 2008 recession \citep{turner2014impact}, and multiple theoretically model university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
%Lastly, instruction costs within individual universities vary substantially across subject areas, primarily thanks to department differences in class size and faculty salaries \citep{hemelt2021math}.
The number of professors per student may be particularly important for quality of instruction and thus educational outcomes;
small class sizes in secondary schools lead to increases in test scores \citep{angrist1999using}, and UK university students perform worse academically in particularly large classes \citep{bandiera2010heterogeneous}.
Though, the economics literature is not decided on whether contingent lecturers lead to better or worse outcomes in higher education.
\textbf{Beef up this part to focus more on negative effects on students.}
Lecturers are more effective teachers than research professors at a highly selective university \citep{bettinger2010does,figlio2015tenure}, but students taught by part-time lecturers have worse student outcomes relative to full-time lecturers or faculty \citep{zhu2021limited,ehrenberg2005tenured,jaeger2011examining}.

\textbf{Tenure research \citep{mcpherson1999tenure,mcpherson1983economics}}

Mine is the frst paper to provide causal evidence of the impact of state funding for higher education on faculty composition and wages.
The closest related research has examined (\textbf{theoretical papers here}) $\hdots$

My results provide suggestive evidence on the mechanisms for previous research on the negative effects of state funding cuts on student outcomes at US public universities \citep{zhu2021limited,ehrenberg2012american}.

Lastly, my results providde evidence that the long term trends in higher education funding and of non-tenure track faculty are causally related.
\textbf{Sentence on how they will continue.}

>>>>

Public universities educate the majority of higher education students in the US, yet have experienced a stagnation in state funding over the last three decades .
At the same time, public university employment of lecturers dramatically increased at public universities, while enrolment rose dramatically.
Previous research has examined the effects of the decline in state funding on student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence of the effects of this funding decline on faculty.
I show that four-year, degree-granting public universities have experienced a systematic decline in state funding per student, and use a shift-share instrument to show that this fall led to a substitution away from tenure-track/tenured professors towards contingent lecturers.
Analysis of all public university faculty in the state of Illinois shows that incumbent faculty were relatively unaffected, in terms of wages, promotion and quit rates, implying that changes in faculty composition arose by disrupting public universities' ability to hire or replace their tenure-track/tenured faculty. 
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors, so that the stagnating state funding for public universities has implications for the wider structure of US higher education and research.

There are various margins through which funding cuts can impact a university's faculty.
A university's level of resources affect the ability that departments can hire new professors, retain their current faculty, or provide retirement incentives.
So that it is natural to expect that funding cuts may affect the composition of professors at a university.
Similarly, the wages a university pays their faculty is affected by funding for the university, including the starting salary it offers to new professors and the amount of yearly raises (if any) for incumbent faculty.
So that it is natural to expect that systematic funding cuts will lead to changes in faculty composition, and possibly affect incumbent professors, too.
Yet, it is a priori unclear whether/how much each of these margins adjusts to funding cuts, as these decision depend on how the university administrations view the productivity of different inputs, plus the degree of wage and employment stickiness over time.

@@

I employ a shift-share instrument to identify changes in state funding for higher education, which exploits how universities differ in their financial reliance on state funding, combined with yearly shocks to the amount an entire state funds higher education --- following \cite{NBERw23736,NBERw27885}.
State governments decide how to fund higher education by a complicated process which can be influenced by local economic conditions, changing state priorities, or even lobbying from the state universities themselves.
As such, state funding is not randomly assigned to universities and so it is necessary to employ an identification strategy, such as the shift-share instrument, to account for this.
The shift-share instrument interacts how much a university relies on state funding as a percent of revenues in a base period with the entire state's total funding per higher education student for each year, exploiting changes in funding and how much each university historically relied on state funding for higher education.

Falls in state funding affect faculty composition at public universities, which substitute away from tenured and tenure-track professors towards contingent lecturers.
A funding cut of \$1,000 per student leads to an average university employing 6 more lecturers;
in percentage terms, a (funding per student) cut of 10\% leads to a fall of 1.4\% in the number of assistant professors per student at a university, a fall of 1.2\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock, showing that the effect is not isolated to the year of the funding shock. 
Over the same time period, state funding per student fell by around 35\%, count of (tenure-tack/tenured) professors per student fell by 9\%, and lecturers per student increased by 99\%, so that these results show that falls in state funding explain around 53\% of the fall in professors per student
% i.e. -(35 * 0.137) / -9
and 15.5\% of the rise in lecturers per student.
% ie., (35 * 0.437) / 99
Additionally, I use individual level data on all Illinois public university professors in 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by the falls in state funding (and by the funding shocks).
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or leaving) professors.

Mine is the first paper to provide causal evidence on the impact of state funding on faculty composition, and wages.
The closest related research on higher education funding has examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,brown2014endowment}.
For faculty outcomes, multiple theoretically model university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}), while others measured how faculty at universities were affected by endowment shocks and the 2008 recession \citep{brown2014endowment,turner2014impact}.
These papers measure effects of funding and changes in university revenues, but not measure effects on faculty, or other outcomes related to university instruction.

My results provide suggestive evidence on mechanisms for the previous research connecting state funding cuts for higher education and worsening student outcomes.
Substitution towards lecturers is likely a cost-cutting measure for funding constrained universities, though relying on these faculty --- who are often over-worked and not granted long-term employment protections --- may lead to worse student outcomes \citep{ehrenberg2005tenured,zhu2021limited,jaeger2011examining}.
Lastly, my results provide evidence that the long term trends in higher education funding are causally related, contributing to the literature in the long run trends in faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing}.

### 2023-07-16 text/sections/04-empirical.tex
writing revision 2 mostly finished.

It is not a priori clear which units are appropriate for this analysis;
does it make sense to consider state funding in purely dollar amounts per student, or in percent change rate?
The funding shock instrument is a strong predictor for the average university's level of state funding in either unit; a funding shock of \$1,000 per student in the entire state leads to around \$1,176 per student at the university, while a funding shock of -10\% leads to around -9.77\% less state funding per year.
Yet, the level of state funding (and the outcome variables) vary greatly between states for the unit of analysis.
For example, the average Illinois public university receives \$10,709 in state funding per student in 1990 and  \$6,713 in 2021 (a fall of over 30\%), while California went from \$19,224 per student to \$12,915 in the same time span (a fall of 37\%).
While most states are not exactly the same to California and Illinois, this is example is telling for the phenomenon of stagnating state funding:
states vary in how much the funding higher education in 1990, but most have experienced a decline on order of 30\%, so that the stagnation in state funding has been a percent change trend over this time period.
As such, I include regression specifications where the explanatory and outcome variables are $\log$ transformed, and refer to these when stating results in percent change terms.

>>>>

It is not a priori clear which units are appropriate for this analysis;
does it make sense to consider state funding in purely dollar amounts per student, or in percent change rate?
The funding shock instrument is a strong predictor for the average university's level of state funding in either unit; a funding shock of \$1,000 per student in the entire state leads to around \$1,176 per student at the university, while a funding shock of $-10$\% leads to around 9.77\% less state funding per year.
Yet, the level of state funding (and the outcome variables) vary greatly between states for the unit of analysis.
For example, the average Illinois public university receives \$10,709 in state funding per student in 1990 and  \$6,713 in 2021 (a fall of over 30\%), while California went from \$19,224 per student to \$12,915 in the same time span (a fall of 37\%).
While most states are not exactly the same to California and Illinois, this is example is telling for the phenomenon of stagnating state funding:
states vary in how much the funding higher education in 1990, but most have experienced a decline on order of 30\%, so that the stagnation in state funding has been a percent change trend over this time period.
As such, I include regression specifications where the explanatory and outcome variables are $\log$ transformed, and refer to these when stating results in percent change terms.
